---
title: "Targeted Learning for Survival Analysis with Competing Risks"
author: "David Benkeser & Nima Hejazi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Targeted Learning for Survival Analysis with Competing Risks}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Introduction

The `survtmle` package is designed to use targeted minimum loss-based estimation
(TMLE) to compute baseline covariate-adjusted estimates of marginal cumulative
incidence in right-censored survival settings with (and without) competing
risks. The package facilitates flexible modeling to adjust for covariates
through the use of ensemble machine learning via the
[`SuperLearner`](https://CRAN.R-project.org/package=SuperLearner)
package.

---

## Single failure type

We examine the use of `survtmle` in a variety of simple examples. The package
can be loaded as follows:

```{r core_pkgs, echo=FALSE}
library(tibble)
```

```{r load_pkg}
library(survtmle)
```

We simulate a simple data with no censoring and a single cause of failure to
illustrate the machinery of the `survtmle` package.

```{r sim_data}
set.seed(1234)
n <- 200
t_0 <- 6
trt <- rbinom(n, 1, 0.5)
adjustVars <- data.frame(W1 = round(runif(n)), W2 = round(runif(n, 0, 2)))
ftime <- round(1 + runif(n, 1, 4) - trt + adjustVars$W1 + adjustVars$W2)
ftype <- round(runif(n, 0, 1))
```

The simple data structure contains a set of baseline covariates (`adjustVars`), 
a binary treatment variable (`trt`), a failure time that is a function of the
treatment, adjustment variables, and a random error (`ftime`), and a failure
type (`ftype`), which denotes the cause of failure (0 means no failure, 1 means
failure). The first few rows of data can be viewed as follows.

```{r view_data, echo=FALSE}
d <- as_data_frame(cbind(ftype, ftime, trt, adjustVars))
d
```

---

### Using the method of iterated means

#### Covariate adjustment via logistic regression

A common goal is to compare the incidence of failure at a fixed time between the
two treatment groups. Covariate adjustment is often desirable in this comparison
to improve efficiency [@moore2009increasing]. This covariate adjustment may be
facilitated by estimating a series of iterated covariate-conditional means
[@robins1999ltmle,@bang2005doubly,@vdlgruber:2012:ijb]. The final iterated
covariate-conditional mean is marginalized over the empirical distribution of
baseline covariates to obtain an estimate of the marginal cumulative incidence. 

Here, we invoke the eponymous `survtmle` function to compute the iterated
mean-based (`method = "mean"`)  covariate-adjusted estimates of the cumulative
incidence at time six (`t0 = 6`) in each of the treatment groups using
quasi-logistic regression (formula specified via `glm.ftime`) to estimate the
iterated means. The `glm.ftime` argument should be a valid right-hand-side
formula specification based on `colnames(adjustVars)` and `"trt"`. Here we use a
simple main terms regression.

```{r example_glm_mean}
# Fit 1: Use GLM-based estimators for failure w/ "mean" method
fit1 <- survtmle(ftime = ftime, ftype = ftype,
                 trt = trt, adjustVars = adjustVars,
                 glm.ftime = "trt + W1 + W2",
                 method = "mean", t0 = t_0)
fit1
```

Internally, `survtmle` estimates the covariate-conditional treatment probability
(via `glm.trt` or `SL.trt`, see below) and covariate-conditional censoring
distribution (via `glm.ctime` or `SL.ctime`, see below). In the above example,
the treatment probability does not depend on covariates (as in e.g., a
randomized trial) and so we did not specify a way to adjust for covariates in
estimating the treatment probability. In this case, `survtmle` sets
`glm.trt = "1"`, which corresponds with empirical estimates of treatment
probability, and sets `glm.ctime` to be equivalent to the Kaplan-Meier censoring
distribution estimates.

In practice, we may wish to adjust for covariates when computing estimates of
the covariate-conditional treatment and censoring probabilities. In
observational studies, the distribution of treatment may differ by measured
covariates, while in almost any study (including randomized trials) it is
possible that censoring differs by covariates. Thus, we often wish to adjust for
covariates to account for measured confounders of treatment receipt and
censoring.

This adjustment may be accomplished using logistic regression through the
`glm.trt` and `glm.ctime` arguments, respectively. The `glm.trt` argument should
be a valid right-hand-side formula specification based on
`colnames(adjustVars)`. The `glm.ctime` argument should be a valid
right-hand-side formula specification based on `colnames(adjustVars)`, `"trt"`,
and `"t"` used to model the hazard function for censoring. By including `"trt"`
and `"t"`, the function allows censoring probabilities to depend on treatment
assignment and time, respectively. Here we call `survtmle` again, now adjusting
for covariates in the treatment and censoring fits.

```{r example_glm_mean_2}
# Fit 2: Use GLM-based estimators for failure, treatment, and censoring with the
#        "mean" method
fit2 <- survtmle(ftime = ftime, ftype = ftype,
                 trt = trt, adjustVars = adjustVars,
                 glm.trt = "W1 + W2",
                 glm.ctime = "W1 + trt + t + I(t^2)",
                 glm.ftime = "trt + W1 + W2",
                 method = "mean", t0 = t_0)
fit2
```

#### Covariate adjustment via Super Learner

While we can certainly use logistic regression to model the treatment,
censoring, and iterated means, a large benefit afforded by the `survtmle`
package is how it leverages `SuperLearner` ensemble machine learning to estimate
these quantities in a more flexible manner. The Super Learner method is a
generalization of stacked regression [@breiman1996stacked] that uses
cross-validation to select the best-performing estimator from a library of
candidate estimators [@vdlpolley:2007:statappgenetics]. Many popular machine
learning algorithms have been implemented in the
[`SuperLearner`](https://cran.r-project.org/package=SuperLearner).

To utilize `SuperLearner` estimates, we can utilize options `SL.trt`,
`SL.ctime`, and `SL.ftime` to estimate conditional treatment, censoring, and
iterated means, respectively. See `?SuperLearner` for details on options for
correctly specifying a super learner library and see `listWrappers()` to print
the methods implemented in the `SuperLearner` package. Here we demonstrate a
call to `survtmle` using a simple library that includes simple algorithms that
are included in base `R`.

```{r example_sl_mean}
# Fit 3: SuperLearner estimators for treatment, failure, and censoring.
fit3 <- survtmle(ftime = ftime, ftype = ftype,
                 trt = trt, adjustVars = adjustVars,
                 SL.trt = c("SL.glm","SL.mean","SL.step"),
                 SL.ftime = c("SL.glm","SL.mean","SL.step"),
                 SL.ctime = c("SL.glm","SL.mean","SL.step"),
                 method = "mean", t0 = t_0)
fit3
```

**Remark:** Invoking `survtmle` with `method = "mean"` and `SL.ftime` requires
fitting a Super Learner *for each time point* from `seq_len(t0)`. If there are
many unique time points observed in the data, this can become a computationally
intensive process. In such cases, we recommend either redefining the `ftime`
variable to pool across time points *or* using `method = "hazard"` (see below). 

---

### Using the method of cause-specific hazards

An alternative method to the iterated mean-based TMLE for estimating cumulative
incidence is based on estimated the (cause-specific) hazard function. This
estimator is implemented by  specifying `method = "hazard"` in a call to
`survtmle`. Just as with `method = "mean"`, we can use either `glm.` or `SL.` to
adjust for covariates. However, now the `glm.ftime` formula may additionally
include functions of time, as this formula is now being used in a pooled
regression to estimate cause-specific hazards over time.

```{r example_glm_hazard}
# Fit 4: GLM estimators for treatment, censoring, and failure w/ "hazard" method
# Note the inclusion of 't' in the formula for glm.ftime.
fit4 <- survtmle(ftime = ftime, ftype = ftype,
                 trt = trt, adjustVars = adjustVars,
                 glm.trt = "W1 + W2",
                 glm.ftime = "trt + W1 + W2 + t + I(t^2)",
                 glm.ctime = "trt + W1 + W2*t",
                 method = "hazard", t0 = t_0)
fit4
```

Here's an example using Super Learner.

```{r example_sl_hazard}
# Fit 5: SuperLearner estimators for failure and censoring, alongside empirical
#        estimators for treatment (the default) using the "hazard" method.
#        Note that the super learner for ftime is also adjusting for time.
fit5 <- survtmle(ftime = ftime, ftype = ftype,
                 trt = trt, adjustVars = adjustVars,
                 SL.trt = c("SL.glm", "SL.mean", "SL.step"),
                 SL.ftime = c("SL.glm", "SL.mean", "SL.step"),
                 SL.ctime = c("SL.glm", "SL.mean", "SL.step"),
                 method = "hazard", t0 = t_0)
fit5
```

**Remark:** The TMLE algorithm for the hazard-based estimator differs from the
iterated mean-based TMLE. In particular, the algorithm is iterative and has no
guarantee of convergence. While we have not identified instances where
convergence is a serious problem, we encourage users to submit any such
situations as [GitHub issues](https://github.com/benkeser/survtmle/issues) or to
write directly to <benkeser@emory.edu>. The stopping criteria for the iteration
may be adjusted via `tol` and `maxIter` options. Increasing `tol` or decreasing
`maxIter` will lead to faster convergence; however, it is recommended that `tol`
be set no larger than `1 / sqrt(length(ftime))`. If `maxIter` is reached without
convergence, one should check that `fit$meanIC` are all less than
`1 / sqrt(length(ftime))`.

---

## Multiple failure types

In all of the preceding examples, we have restricted our attention to the case
where there is only a single failure type of interest. Now we consider more
scenarios where we observe multiple failure types. First, we simulate data with
two types of failure.

```{r simulate_mult_data}
set.seed(1234)
n <- 200
trt <- rbinom(n, 1, 0.5)
adjustVars <- data.frame(W1 = round(runif(n)), W2 = round(runif(n, 0, 2)))
ftime <- round(1 + runif(n, 1, 4) - trt + adjustVars$W1 + adjustVars$W2)
ftype <- round(runif(n, 0, 2))
```

This simulated data structure is similar to the single failure type data;
however, now the failure type variable (`ftype`) now contains two distinct types
of failure (with 0 still reserved for no failure).

```{r examine_mult_data, echo=FALSE}
dm <- tibble::as_tibble(cbind(ftype, ftime, trt, adjustVars))
dm
```

When multiple failure types are present, a common goal is to compare the
cumulative incidence of a particular failure type at a fixed time between the
two treatment groups, while accounting for the fact that participants may fail
due to other failure types. Covariate adjustment is again desirable to improve
efficiency and account for measured confounders of treatment and censoring.

### Using the method of iterated means

#### Covariate adjustment via logistic regression

The call to invoke `survtmle` is exactly the same as in the single failure type
case.

```{r example_glm_mean_multftypes}
# Fit 6: GLM estimators for treatment, censoring, and failure w/ "mean" method
fit6 <- survtmle(ftime = ftime, ftype = ftype,
                 trt = trt, adjustVars = adjustVars,
                 glm.trt = "W1 + W2",
                 glm.ftime = "trt + W1 + W2",
                 glm.ctime = "trt + W1 + W2",
                 method = "mean", t0 = t_0)
fit6
```

The output object contains cumulative incidence estimates for each of the four
groups defined by the two failure types and treatments.

There are sometimes failure types that are not of direct interest to out study.
Because `survtmle` invoked with `method = "mean"` computes an estimate of the
cumulative incidence of each failure type separately, we can save on computation
time by specifying which failure types we care about via the `ftypeOfInterest`
option.

```{r example_glm_mean_multftypes_ftypeOfInterest}
# Fit 7: GLM estimators for treatment, censoring, and failure w/ "mean" method
fit7 <- survtmle(ftime = ftime, ftype = ftype,
                 trt = trt, adjustVars = adjustVars,
                 glm.trt = "W1 + W2",
                 glm.ftime = "trt + W1 + W2",
                 glm.ctime = "trt + W1 + W2",
                 method = "mean", t0 = t_0,
                 ftypeOfInterest = 1)
fit7
```

#### Covariate adjustment via Super Learner

As before, we can use the `SuperLearner` ensemble learning algorithm to adjust
for covariates in multiple failure type settings as well.

```{r example_sl_mean_multftypes}
# Fit 8: SuperLearner estimators for failure and censoring and empirical
#        estimators for treatment (default) using the "mean" method.
fit8 <- survtmle(ftime = ftime, ftype = ftype,
                 trt = trt, adjustVars = adjustVars,
                 SL.trt = c("SL.glm","SL.mean","SL.step"),
                 SL.ftime = c("SL.glm","SL.mean","SL.step"),
                 SL.ctime = c("SL.glm","SL.mean","SL.step"),
                 method = "mean", t0 = t_0)
fit8
```

**Remark:** As with single failure type, the `method = "mean"` call to
`survtmle` may be computationally intensive with many time points. This is
especially true when there are additionally multiple failure types, as the
function must repeat these calls to `SuperLearner` separately for each type of
failure. In this case, calls to `survtmle` could be parallelized with one call
to `survtmle` for each type of failure specifying `ftypeOfInterest`.

### Using the method of cause-specific hazards

#### Covariate adjustment via logistic regression

The TMLE based on cause-specific hazards can also be used to compute cumulative
incidence estimates in settings with multiple failure types. As above, the
`glm.ftime` formula may additionally include functions of time, as this formula
is now being used in a pooled regression to estimate cause-specific hazard of
each failure type over time.

```{r example_glm_hazard_multftypes}
# Fit 9: same as Fit 8 above, but using the "hazard" method
fit9 <- survtmle(ftime = ftime, ftype = ftype,
                 trt = trt, adjustVars = adjustVars,
                 glm.trt = "W1 + W2",
                 glm.ftime = "trt + W1 + W2",
                 glm.ctime = "trt + W1 + W2",
                 method = "hazard", t0 = t_0)
fit9
```

We can also leverage the `SuperLearner` algorithm when using the method of
cause-specific hazards with multiple failure types of interest.

```{r example_sl_hazard_multftypes}
# Fit 10: same as Fit 7 above, but using the "hazard" method
fit10 <- survtmle(ftime = ftime, ftype = ftype,
                  trt = trt, adjustVars = adjustVars,
                  SL.trt = c("SL.glm","SL.mean","SL.step"),
                  SL.ftime = c("SL.glm","SL.mean","SL.step"),
                  SL.ctime = c("SL.glm","SL.mean","SL.step"),
                  method = "hazard", t0 = t_0)
fit10
```

As with the iterated-mean based TMLE, we can obtain estimates of cumulative
incidence of only certain failure types (via `ftypeOfInterest`); however, this
does not necessarily result in faster computation, as it did in the case above.
In situations where the convergence of the algorithm is an issue, it may be
useful to invoke multiple calls to `survtmle` with singular `ftypeOfInterest`.
If such convergence issues arise, please report them as [GitHub
issues](https://github.com/benkeser/survtmle/issues) or contact us at
<benkeser@emory.edu>.


## Estimation in bounded models

In certain situations, we have knowledge that the incidence of an event is
bounded below/above for every strata in the population. It is possible to
incorporate these bounds into the TMLE estimation procedure to ensure that any
resulting estimate of cumulative incidence is compatible with these bounds.
Please refer to @benkeser2017improved for more on bounded TMLEs and their
potential benefits.

Bounds can be passed to `survtmle` by creating a `data.frame` that contains
columns with specific names. In particular, there should be a column named
`"t"`. There should additionally be columns for the lower and upper bound for
each type of failure. For example if there is only one type of failure
(`ftype = 1` or `ftype = 0`) then the bounds `data.frame` can contain columns
`"l1"`, and `"u1"` denote the lower and upper bounds, respectively, on the
iterated conditional mean (for `method = "mean"`) or the conditional hazard
function (for `method = "hazard"`). If there are two types of failure
(`ftype = 1`, `ftype = 2`, or `ftype = 0`) then there can additionally be
columns `"l2"` and `"u2"` denoting the lower and upper bounds, respectively, on
the iterated conditional mean for type two failures (for `method = "mean"`) or
the conditional cause-specific hazard function for type two failures (for
`method = "hazard"`).

Here is a simple example.

```{r bounded_tmle}
bf1 <- data.frame(t = seq_len(t_0), l1 = rep(0.01, t_0), u1 = rep(0.99, t_0))
bf1
```

Now that we have specified our bounds, we can invoke `survtmle` repeating our
first example ("Fit 1"), but now restricting the iterated conditional means to
follow the bounds specified above.

```{r example_glm_mean_bounded}
# Fit 11: Fit 2, but now specifying bounds on the iterated conditional means
fit11 <- survtmle(ftime = ftime, ftype = ftype,
                 trt = trt, adjustVars = adjustVars,
                 glm.trt = "W1 + W2",
                 glm.ftime = "trt + W1 + W2",
                 glm.ctime = "trt + W1 + W2",
                 method = "mean", t0 = t_0,
                 bounds = bf1)
fit11
```

When there are multiple failure types of interest, we can still provide bounds
for the iterated conditional means (or the conditional hazard function,
whichever is appropriate based on our specification of the `method` argument).

```{r}
# need to make a data.frame of bounds in proper format two types of failure that
# are labeled with ftype = 1 and ftype = 2, so bounds should have columns 't',
# 'l1', 'u1', 'l2', and 'u2'.
bf2 <- data.frame(t = seq_len(t_0),
                  l1 = rep(0.01, t_0), u1 = rep(0.99, t_0),
                  l2 = rep(0.02, t_0), u2 = rep(0.99, t_0)
                 )
bf2
```

Now, we invoke `survtmle`, passing in the specified bounds using the appropriate
argument:

```{r}
# Fit 12: same as Fit 5 above, but now include bounds
fit12 <- survtmle(ftime = ftime, ftype = ftype,
                  trt = trt, adjustVars = adjustVars,
                  glm.trt = "W1 + W2",
                  glm.ftime = "trt + W1 + W2",
                  glm.ctime = "trt + W1 + W2",
                  method = "mean", t0 = t_0,
                  bounds = bf2)
fit12
```

**Remark 1:** Please see the discussion in @benkeser2017improved on how to
select bounds for these procedures. Note that poorly chosen bounds can lead to
instability in the estimation procedure.

**Remark 2:** While it is theoretically possible to use super learner to perform
bounded estimation, many of the implemented algorithms are not currently
designed to respect bounds. Nevertheless, it is possible to write one's own
algorithms to incorporate such bounds. However, for the sake of stability, we
have restricted the bounded implementation to `glm` based covariate-adjustment. 

---

## Utility functions

### Estimating cumulative incidence curves

The `survtmle` function provides the function `timepoints` to compute the
estimated cumulative incidence over multiple timepoints. This function is
invoked after an initial call to `survtmle` with option `returnModels = TRUE`.
By setting this option, the `timepoints` function is able to recycle fits for
the conditional treatment probability, censoring distribution, and, in the case
of `method = "hazard"`, the hazard fits. Thus, invoking `timepoints` is faster
than making repeated calls to `survtmle` with different `t0`.

There is some subtlety involved to properly leveraging this facility. Recall
that the censoring distribution fit (and cause-specific hazard fit) pools over
all time points. Thus, in order to most efficiently use `timepoints`, the
initial call to `survtmle` should be made setting option `t0` equal to the final
time point at which one wants estimates of cumulative incidence. This allows
these hazard fitting procedures to utilize all of the data to estimate the
conditional hazard function.

We demonstrate the use of `timepoints` below based on the following simulated
data.

```{r sim_data_2}
set.seed(1234)
n <- 200
t_0 <- 6
trt <- rbinom(n, 1, 0.5)
adjustVars <- data.frame(W1 = round(runif(n)), W2 = round(runif(n, 0, 2)))
ftime <- round(1 + runif(n, 1, 4) - trt + adjustVars$W1 + adjustVars$W2)
ftype <- round(runif(n, 0, 1))
```

Imagine that we would like cumulative incidence estimates at times
`seq_len(t_0)` based on `fit2` above (mean-based TMLE using glm covariate
adjustment). However, note that when we originally called `fit2` the option
`returnModels` was set to its default value `FALSE`. Thus, we must refit this
object setting the function to return the model fits.

```{r example_glm_mean_3}
# Refit fit 2 returning models
fit2_rm <- survtmle(ftime = ftime, ftype = ftype,
                 trt = trt, adjustVars = adjustVars,
                 glm.trt = "W1 + W2",
                 glm.ctime = "W1 + trt + t + I(t^2)",
                 glm.ftime = "trt + W1 + W2",
                 method = "mean", t0 = t_0,
                 returnModels = TRUE)
fit2_rm
```

Now we can call `timepoints` to return estimates of cumulative incidence at each
time `seq_len(t_0)`.

```{r}
tp.fit2 <- timepoints(fit2_rm, times = seq_len(t_0))
# print the object
tp.fit2
```

Internally, `timepoints` is making calls to `survtmle`, but is passing in the
fitted treatment and censoring fits from `fit2_rm$trtMod` and
`fit2_rm$ctimeMod`. However, for `method = "mean"` the function is still fitting
the iterated means separately for each time required by the call to
`timepoints`. Thus, the call to `timepoints` may be quite slow if
`method = "mean"`, `SL.ftime` is specified (as opposed to `glm.ftime`), and/or
many times are passed in via `times`. Future implementations may attempt to
avoid this extra model fitting. For now, if many times are required, we
recommend using `method = "hazard"`, which is able to recycle all of the model
fits. Below is an example of this.

```{r}
# Refit Fit 4, setting returnModels=TRUE this time...
fit4_rm <- survtmle(ftime = ftime, ftype = ftype,
                 trt = trt, adjustVars = adjustVars,
                 glm.trt = "W1 + W2",
                 glm.ftime = "trt + W1 + W2 + t + I(t^2)",
                 glm.ctime = "trt + W1 + W2*t",
                 method = "hazard", t0 = t_0,
                 returnModels = TRUE)
# call timepoints based on this fit
tp.fit4 <- timepoints(fit4_rm, times = seq_len(t_0))
# print the object
tp.fit4
```

There is a plotting method available for `timepoints` to plot cumulative
incidence over time in each treatment group and for each failure type.

```{r}
# plot raw cumulative incidence
plot(tp.fit4, type = "raw")
```

Because the cumulative incidence function is being invoked pointwise, it is
possible that the resulting curve is not monotone. However, it is possible to
show that projecting this curve onto a monotone function via isotonic regression
results in an estimate with identical asymptotic properties to the pointwise
estimate. Therefore, we additionally provide an option `type = "iso"` (the
default) that provides these smoothed curves.

```{r}
# plot smoothed cumulative incidence
plot(tp.fit4)
```
---

## Session Information

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

---

## References
